---
title: "Bus Headway Analysis: Route 1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bus Headway Analysis: Route 1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE  # set to TRUE locally to hit the API
)
```

Route 1 is one of the busiest bus routes in the MBTA system, running between
Harvard Square and Nubian Station through Cambridge and the South End. This
vignette uses the stop lookup functions to pull and plot headway data without
needing to know a single stop ID up front.

## Setup

```{r load}
library(transitmattr)
library(dplyr)
library(ggplot2)
```

## Step 1: Discover the route

Start by listing all stops on Route 1 so you can pick the ones you care about.

```{r stations}
tm_bus_stations("1")
```

You'll see stops listed in order from Harvard to Nubian. The `station` column
holds the short code used internally; `stop_name` is what you'll use with
`tm_bus_stop_id()`.

Since bus directions are always `"outbound"` (away from downtown) and
`"inbound"` (toward downtown), we don't need a discovery step for directions.

## Step 2: Pull headways for a single day

Let's look at headways at **MIT @ Mass Ave** — a busy mid-route stop — for both
directions on a typical winter weekday.

```{r pull-headways}
date <- "2024-01-16"  # Tuesday

inbound  <- tm_headways(date,
                        stop = tm_bus_stop_id("1", "MIT @ Mass Ave", "inbound"))
outbound <- tm_headways(date,
                        stop = tm_bus_stop_id("1", "MIT @ Mass Ave", "outbound"))
```

Each result is a list of individual bus arrivals. Bind them into a single data
frame and add a `direction` label:

```{r bind-headways}
headways_df <- bind_rows(
  bind_rows(inbound)  |> mutate(direction = "Inbound (→ Nubian)"),
  bind_rows(outbound) |> mutate(direction = "Outbound (→ Harvard)")
)

glimpse(headways_df)
```

## Step 3: Convert times and filter

The departure timestamp (`current_dep_dt`) comes back as an ISO 8601 string
(e.g. `"2024-01-16T09:30:00"`). Convert it and keep only the daytime service
window.

```{r clean}
headways_df <- headways_df |>
  mutate(
    dep_time      = as.POSIXct(current_dep_dt, format = "%Y-%m-%dT%H:%M:%S",
                                tz = "America/New_York"),
    headway_min   = headway_time_sec / 60,
    benchmark_min = benchmark_headway_time_sec / 60,
    hour          = as.integer(format(dep_time, "%H"))
  ) |>
  filter(hour >= 6, hour <= 22)
```

## Step 4: Plot headways over the day

```{r plot-scatter}
ggplot(headways_df, aes(x = dep_time, y = headway_min, color = direction)) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  scale_color_manual(values = c(
    "Inbound (→ Nubian)"    = "#003DA5",
    "Outbound (→ Harvard)"  = "#80A8DD"
  )) +
  labs(
    title    = "Route 1 Headways at MIT @ Mass Ave",
    subtitle = paste("January 16, 2024 ·", date),
    x        = NULL,
    y        = "Minutes between buses",
    color    = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

> **What to look for:** Peaks in headway mean longer waits between buses — often
> during the AM peak (more inbound demand) or the PM peak (more outbound demand).
> A headway above 15 minutes on a route that's supposed to run every 8 minutes
> suggests a gap in service.

## Step 5: Compare actual vs. scheduled headway

The `benchmark_headway_time_sec` field is what the schedule promised. Comparing
it to actual headway shows where the route is running behind schedule.

```{r plot-vs-benchmark}
headways_df |>
  filter(direction == "Inbound (→ Nubian)") |>
  mutate(
    excess_min = headway_min - benchmark_min
  ) |>
  ggplot(aes(x = dep_time, y = excess_min)) +
  geom_col(aes(fill = excess_min > 0), width = 60) +
    scale_fill_manual(
      values = c("FALSE" = "#80A8DD", "TRUE" = "#DA291C"),
      labels = c("Under schedule", "Over schedule"),
      name   = NULL
    ) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    labs(
      title    = "Route 1 Inbound — Excess Wait vs. Schedule",
      subtitle = "MIT @ Mass Ave · January 16, 2024",
      x        = NULL,
      y        = "Extra minutes beyond scheduled headway"
    ) +
    theme_minimal() +
    theme(legend.position = "top")
```

> **Red bars** are moments when the actual gap between buses exceeded the
> scheduled gap. Clusters of red bars often indicate bus bunching — a common
> problem on long surface routes where small delays compound.

## Step 6: Aggregate over a month

For a broader view, use `tm_aggregate_headways()`, which returns daily averages
rather than individual events.

```{r aggregate}
agg_raw <- tm_aggregate_headways(
  stop       = tm_bus_stop_id("1", "MIT @ Mass Ave", "inbound"),
  start_date = "2024-01-01",
  end_date   = "2024-01-31"
)

agg_df <- bind_rows(agg_raw) |>
  mutate(date = as.Date(service_date))
```

```{r plot-aggregate}
ggplot(agg_df, aes(x = date, y = mean / 60)) +
  geom_line(color = "#003DA5") +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed",
              color = "grey50") +
  labs(
    title    = "Route 1 Average Inbound Headway — January 2024",
    subtitle = "MIT @ Mass Ave",
    x        = NULL,
    y        = "Average headway (minutes)"
  ) +
  theme_minimal()
```

## Going further

Now that you can look up stops by name, try exploring other routes and stops:

```{r explore, eval = FALSE}
# Which bus routes are available?
tm_bus_routes()

# Stops on Route 66 (Harvard → Dudley)
tm_bus_stations("66")

# Headways at Nubian on Route 1 — the southern terminus
tm_headways("2024-01-16",
  stop = tm_bus_stop_id("1", "Nubian Station", "inbound")
)
```

See the **Finding Stops by Name** vignette for the full lookup API across all
transit modes, including commuter rail and ferry.
