---
title: "Finding Stops by Name"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding Stops by Name}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r load, include = FALSE}
library(transitmattr)
```

In `transitmattr`, you never have to look up or memorise stop IDs. Instead,
you build a **query object** for a transit mode and line/route, then call
`$stop("Station Name")` — the library resolves the correct stop ID
automatically.

This vignette shows how to discover valid station names and build queries for
all four transit modes.

---

## Rapid Transit

The five rapid transit lines are **Red**, **Orange**, **Blue**, **Green**, and
**Mattapan**.

### Discover: what stations are on this line?

```{r rt-stations}
tm_rt_query("Red")$stations()
```

### Discover: which directions does this line use?

Rapid transit lines use geographic direction names rather than inbound/outbound.

```{r rt-directions}
tm_rt_query("Red")$directions()   # northbound / southbound
tm_rt_query("Green")$directions() # eastbound  / westbound
```

### Build a query and fetch data

```{r rt-api, eval = FALSE}
# Headways at Davis, all trains (direction resolved from stop ID)
tm_rt_query("Red")$stop("Davis")$headways("2024-01-15")

# For travel times, specify direction (determines which platform stop ID to use)
tm_rt_query("Red")$from_stop("Park Street")$to_stop("Davis")$direction("northbound")$travel_times("2024-01-15")

# Aggregate headways for January 2024 (direction filters to one platform)
tm_rt_query("Red")$stop("Davis")$direction("southbound")$date_range("2024-01-01", "2024-01-31")$aggregate_headways()
```

You can chain builder calls before the terminal call. The query object accumulates
context and validates each input as you set it — typos in station names produce
an error immediately, not at request time.

---

## Bus

There are 121 bus routes. Directions are always `"outbound"` and `"inbound"`.

### Discover: what routes are available?

```{r bus-routes}
head(tm_bus_routes(), 15)
```

### Discover: what stops are on this route?

```{r bus-stations}
tm_bus_query("1")$stations()
```

### Build a query and fetch data

```{r bus-api, eval = FALSE}
# Headways at Harvard on Route 1, inbound, on a specific day
tm_bus_query("1")$stop("Harvard")$headways("2024-01-15")

# Aggregate headways at Harvard over a month
tm_bus_query("1")$stop("Harvard")$direction("inbound")$date_range("2024-01-01", "2024-01-31")$aggregate_headways()
```

---

## Commuter Rail

Commuter Rail routes are identified by their GTFS route ID. All 13 lines are
available.

### Discover: what routes are available?

```{r cr-routes}
tm_cr_routes()
```

### Discover: what stops are on this line?

The `terminus` column flags the endpoints of each line. Terminus stations have
stop IDs for only one direction — calling `$stop()` on a terminus station will
issue a **warning** to alert you that data may be incomplete in one direction.

```{r cr-stations}
tm_cr_query("CR-Fairmount")$stations()
```

### Build a query and fetch data

```{r cr-api, eval = FALSE}
# Headways at Fairmount, inbound trains (toward South Station)
tm_cr_query("CR-Fairmount")$stop("Fairmount")$headways("2026-06-18")

# Travel time, Fairmount → Blue Hill Avenue, inbound
tm_cr_query("CR-Fairmount")$from_stop("Fairmount")$to_stop("Blue Hill Avenue")$direction("inbound")$travel_times("2026-06-18")
```

> **Terminus warning:** If you call `$stop("South Station")` on a CR query, you
> will see a warning — South Station is a terminus and has stop IDs in only one
> direction. The query still works; the warning is informational.

---

## Ferry

Six ferry routes are available. Like bus and CR, directions are `"outbound"`
and `"inbound"`.

### Discover: what routes are available?

```{r ferry-routes}
tm_ferry_routes()
```

### Discover: what stops are on this route?

```{r ferry-stations}
tm_ferry_query("Boat-F1")$stations()
```

### Build a query and fetch data

```{r ferry-api, eval = FALSE}
# Headways at Hingham, inbound (toward Long Wharf)
tm_ferry_query("Boat-F1")$stop("Hingham")$headways("2024-06-01")
```

---

## Quick reference

| Mode | Routes fn | Query constructor | Stations method |
|------|-----------|-------------------|-----------------|
| Rapid transit | — | `tm_rt_query(line)` | `$stations()` |
| Bus | `tm_bus_routes()` | `tm_bus_query(route)` | `$stations()` |
| Commuter Rail | `tm_cr_routes()` | `tm_cr_query(route)` | `$stations()` |
| Ferry | `tm_ferry_routes()` | `tm_ferry_query(route)` | `$stations()` |

All station name arguments are **case-insensitive**. Line and route arguments
accept multiple forms (`"red"`, `"Red"`, and `"line-Red"` are all equivalent).
