---
title: "Finding Stops by Name"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding Stops by Name}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r load, include = FALSE}
library(transitmattr)
```

Every API function that returns per-trip data (headways, dwells, travel times)
needs a **stop ID** — a string that identifies exactly where on the network you
want to look. These IDs are opaque by design: `"70061"` is Alewife, `"1-1-110"`
is Harvard inbound on Route 1, `"Boat-F1|1|Boat-Hingham"` is Hingham Ferry
inbound. Nobody has these memorised.

The `tm_*_stop_id()` family converts human-readable names into the IDs the API
expects. This vignette is a quick reference for all four transit modes.

---

## Rapid Transit

The five rapid transit lines are **Red**, **Orange**, **Blue**, **Green**, and
**Mattapan**.

### Discover: what stations are on this line?

```{r rt-stations}
tm_line_stations("Red")
```

### Discover: which directions does this line use?

Rapid transit lines use geographic direction names rather than inbound/outbound.

```{r rt-directions}
tm_line_directions("Red")   # northbound / southbound
tm_line_directions("Green") # eastbound  / westbound
```

### Look up a stop ID

```{r rt-stop-id}
# Davis, trains heading toward Alewife (northbound)
tm_stop_id("Red", "Davis", "northbound")

# Alewife is a terminus — same stop ID in both directions
tm_stop_id("Red", "Alewife", "northbound")
tm_stop_id("Red", "Alewife", "southbound")
```

You can also use `0`/`1` instead of direction names if you prefer:

```{r rt-stop-id-numeric}
tm_stop_id("Red", "Davis", 0)  # same as "northbound"
tm_stop_id("Red", "Davis", 1)  # same as "southbound"
```

### Look up a place ID

`tm_place_id()` returns the GTFS parent-station identifier for a rapid transit
station (e.g. `"place-pktrm"`). These are **not** used by the TransitMatters
aggregate endpoints — they are provided as a convenience if you need to cross-
reference with the MBTA v3 API.

```{r rt-place-id}
tm_place_id("Red", "Park Street")
tm_place_id("Red", "Davis")
```

### Using these in API calls

```{r rt-api}
# Headways at Davis, trains heading toward Alewife
davis_headways = tm_headways("2024-01-15",
  stop = tm_stop_id("Red", "Davis", "northbound")
)

# Travel time: Park Street → Davis (aggregate endpoint uses stop IDs)
parkst_2_davis_tt = tm_aggregate_travel_times2(
  from_stop  = tm_stop_id("Red", "Park Street", "northbound"),
  to_stop    = tm_stop_id("Red", "Davis", "northbound"),
  start_date = "2024-01-01",
  end_date   = "2024-01-31"
)

```

---

## Bus

There are 121 bus routes in the constants. Directions are always
`"outbound"` and `"inbound"`.

### Discover: what routes are available?

```{r bus-routes}
# Returns all route IDs — sorted, so standard routes come first
head(tm_bus_routes(), 15)
```

### Discover: what stops are on this route?

```{r bus-stations}
# Route 1 runs between Harvard and Nubian
tm_bus_stations("1")
```

### Look up a stop ID

```{r bus-stop-id}
# Harvard, boarding a Route 1 toward Nubian (inbound)
tm_bus_stop_id("1", "Harvard", "inbound")

# Harvard, boarding a Route 1 toward Watertown (outbound)
tm_bus_stop_id("1", "Harvard", "outbound")
```

Bus stop IDs encode the route and direction — `"1-1-110"` means route 1,
direction 1 (inbound), stop sequence 110.

### Using these in API calls

```{r bus-api}
# Headways at Harvard on Route 1, inbound, on a specific day
harvard_headways = tm_headways("2024-01-15",
  stop = tm_bus_stop_id("1", "Harvard", "inbound")
)

# Aggregate headways at Nubian over a month
aggregate_headways_1 = tm_aggregate_headways(
  stop       = tm_bus_stop_id("1", "Nubian Station", "inbound"),
  start_date = "2024-01-01",
  end_date   = "2024-01-31"
)
```

---

## Commuter Rail

Commuter Rail routes are identified by their GTFS route ID. All 13 lines are
available.

### Discover: what routes are available?

```{r cr-routes}
tm_cr_routes()
```

### Discover: what stops are on this line?

The `terminus` column flags the endpoints of each line.

```{r cr-stations}
tm_cr_stations("CR-Fairmount")
```

### Look up a stop ID

CR stations can have multiple platform stop IDs for the same station. At
**terminus** stations, one direction may have no stops (trains only depart, not
arrive, in the other direction).

```{r cr-stop-id}
# Fairmount station, outbound (toward Readville)
tm_cr_stop_id("CR-Fairmount", "Fairmount", "outbound")

# South Station, inbound — terminus, so no inbound stops
tm_cr_stop_id("CR-Fairmount", "South Station", "inbound")

# South Station, outbound — all platforms
tm_cr_stop_id("CR-Fairmount", "South Station", "outbound")
```

### Look up a place ID

`tm_cr_place_id()` returns the GTFS parent-station identifier for a CR station
(e.g. `"place-sstat"`). Like `tm_place_id()` for rapid transit, these are **not**
used by the TransitMatters aggregate endpoints — they are provided for cross-
referencing with the MBTA v3 API.

```{r cr-place-id}
tm_cr_place_id("CR-Fairmount", "South Station")
tm_cr_place_id("CR-Fairmount", "Fairmount")
```

### Using these in API calls

```{r cr-api}
# Headways at Fairmount station, outbound
tm_headways("2024-01-15",
  stop = tm_cr_stop_id("CR-Fairmount", "Fairmount", "outbound")
)

# Aggregate travel time, Fairmount → South Station (pass all platform stop IDs)
tm_aggregate_travel_times(
  from_stop  = tm_cr_stop_id("CR-Fairmount", "Fairmount", "outbound"),
  to_stop    = tm_cr_stop_id("CR-Fairmount", "South Station", "outbound"),
  start_date = "2024-01-01",
  end_date   = "2024-01-31"
)
```

---

## Ferry

Six ferry routes are available. Like bus and CR, directions are `"outbound"`
and `"inbound"`.

### Discover: what routes are available?

```{r ferry-routes}
tm_ferry_routes()
```

### Discover: what stops are on this route?

```{r ferry-stations}
# F1: Long Wharf ↔ Hingham
tm_ferry_stations("Boat-F1")
```

### Look up a stop ID

```{r ferry-stop-id}
# Hingham, boarding a ferry to Long Wharf (inbound)
tm_ferry_stop_id("Boat-F1", "Hingham", "inbound")

# Long Wharf (North), boarding a ferry to Hingham (outbound)
tm_ferry_stop_id("Boat-F1", "Long Wharf (North)", "outbound")
```

### Using these in API calls

```{r ferry-api}
# Headways at Hingham, inbound
ferry_headways = tm_headways("2024-06-01",
  stop = tm_ferry_stop_id("Boat-F1", "Hingham", "inbound")
)
```

---

## Quick reference

| Mode | Routes fn | Stations fn | Stop ID fn | Place ID fn |
|------|-----------|-------------|------------|-------------|
| Rapid transit | — | `tm_line_stations(line)` | `tm_stop_id(line, stop, dir)` | `tm_place_id(line, stop)` |
| Bus | `tm_bus_routes()` | `tm_bus_stations(route)` | `tm_bus_stop_id(route, stop, dir)` | — |
| Commuter Rail | `tm_cr_routes()` | `tm_cr_stations(route)` | `tm_cr_stop_id(route, stop, dir)` | `tm_cr_place_id(route, stop)` |
| Ferry | `tm_ferry_routes()` | `tm_ferry_stations(route)` | `tm_ferry_stop_id(route, stop, dir)` | — |

All stop-name arguments are **case-insensitive**. Line and route arguments are
also case-insensitive (`"red"` and `"Red"` both work).
