---
title: "Getting Started with transitmattr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with transitmattr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## What is this package?

`transitmattr` lets you pull live and historical data from the
[TransitMatters Data Dashboard API](https://dashboard.transitmatters.org)
directly into R. You can get things like:

- Which routes and stops are available
- How often trains are coming (headways)
- How long trains sit at stations (dwell times)
- How long it takes to travel between stops (travel times)
- How many people are riding (ridership)
- Where slow zones are affecting service

## Installation

```{r install, eval = FALSE}
# install.packages("pak")  # run this once if you don't have pak
pak::pak("transitmatters/transitmattr")
```

## Loading the package

```{r load}
library(transitmattr)
library(dplyr)
library(ggplot2)
```

## Step 1: Check that the API is working

Before doing anything else, it's good practice to confirm the API is up.

```{r healthcheck, eval = FALSE}
tm_healthcheck()
```

You'll get back something like:

```
$status
[1] "pass"
```

If you see `"pass"` you're good to go.

## Step 2: Understanding what you get back

All `tm_*` functions return a **list** — R's way of bundling different pieces
of data together. Most dated endpoints return an unnamed list of event records,
one per observation (e.g. one per train departure). Use `str()` to see the
shape and `[[1]]` to inspect a single record.

```{r list-example, eval = FALSE}
result <- tm_headways("2024-01-15", stop = "70061")

length(result)    # number of events
str(result[[1]])  # fields: route_id, direction, current_dep_dt,
                  #   headway_time_sec, benchmark_headway_time_sec,
                  #   vehicle_label, vehicle_consist
```

## Step 3: Discover routes and stops

Before pulling event data you often need to know which routes and stop IDs exist.

```{r routes-stops, eval = FALSE}
# All available routes, grouped by mode
routes <- tm_routes()
names(routes)  # "rapid_transit" "bus" "commuter_rail" "ferry"

# Stop IDs for a specific route (needed for filtering headways, travel times, etc.)
# Route IDs come from tm_routes() — e.g. "Red", "Orange", "Green-B"
red_stops <- tm_stops("Red")
```

## Step 4: Pull data for a single day

These functions need a date. You can also pass stop IDs to narrow the results —
without them you get every stop, which can be a large response.

```{r single-day, eval = FALSE}
# Headways at one stop (or a vector of stops)
headways <- tm_headways("2024-01-15", stop = "70061")
headways <- tm_headways("2024-01-15", stop = c("70061", "70063"))

# Dwell times at a stop
dwells <- tm_dwells("2024-01-15", stop = "70061")

# Travel times between a stop pair
travel <- tm_travel_times("2024-01-15", from_stop = "70076", to_stop = "70064")

# Alerts for a specific day and route
alerts <- tm_alerts("2024-01-15", route = "Red")

# Current live alerts for a route
alerts_now <- tm_alerts(route = "Red")
```

## Step 5: Pull data over a date range

Most analysis requires looking at a range of dates. These functions take
`start_date` and `end_date` plus some extra options.

```{r range, eval = FALSE}
# Daily Red Line ridership for January 2024
ridership <- tm_ridership(
  start_date = "2024-01-01",
  end_date   = "2024-01-31",
  line_id    = "line-red"
)

# Daily trip metrics for the Red Line
trips <- tm_trip_metrics(
  start_date = "2024-01-01",
  end_date   = "2024-01-31",
  agg        = "daily",
  line       = "line-red"
)
```

**Available line IDs:**

| Line        | `line_id` / `line`  |
|-------------|---------------------|
| Red Line    | `"line-red"`        |
| Orange Line | `"line-orange"`     |
| Blue Line   | `"line-blue"`       |
| Green Line  | `"line-green"`      |
| Commuter    | `"line-cr"`         |

## Step 6: Turn the result into a data frame

The API gives back a list, but most R analysis tools (like `ggplot2`) prefer a
**data frame**. Here's the easiest way to convert:

```{r to-df, eval = FALSE}
library(dplyr)
library(transitmattr)

# Pull ridership data
ridership_raw <- tm_ridership("2024-01-01", "2024-01-31", line_id = "line-red")

# Each item in ridership_raw is one record — bind them into rows
ridership_df <- bind_rows(ridership_raw)

# Now it looks like a spreadsheet
head(ridership_df)
```

> **Tip:** If `bind_rows()` doesn't work, try
> `do.call(rbind, lapply(ridership_raw, as.data.frame))` — it's a bit more
> verbose but handles some tricky list shapes.

## Step 7: Make a quick plot

Once you have a data frame you can plot with `ggplot2`:

```{r plot, eval = FALSE}
library(ggplot2)
library(dplyr)
library(transitmattr)

ridership_df <- bind_rows(
  tm_ridership("2024-01-01", "2024-01-31", line_id = "line-red")
)

ggplot(ridership_df, aes(x = as.Date(date), y = count)) +
  geom_line(color = "#DA291C") +   # Red Line color
  labs(
    title = "Red Line Daily Ridership — January 2024",
    x     = "Date",
    y     = "Riders"
  ) +
  theme_minimal()
```

## Recap: the functions at a glance

| Function | What it returns | Needs |
|---|---|---|
| `tm_healthcheck()` | API status | nothing |
| `tm_facilities()` | All MBTA stations | nothing |
| `tm_routes()` | All available routes by mode | nothing |
| `tm_stops(route_id)` | Stop IDs for a route | route ID |
| `tm_headways(date, stop)` | Train frequency | date + stop(s) |
| `tm_dwells(date, stop)` | Station dwell times | date + stop(s) |
| `tm_travel_times(date, from_stop, to_stop)` | Stop-to-stop travel | date + stop pair |
| `tm_alerts(date, route)` | Service alerts | optional date + route |
| `tm_ridership(start, end)` | Ridership counts | date range |
| `tm_trip_metrics(start, end, agg, line)` | Trip performance | date range + line |
| `tm_line_delays(start, end, line)` | Delay summaries | date range + line |
| `tm_speed_restrictions(line_id, date)` | Slow zones | line + date |
| `tm_aggregate_travel_times(...)` | Long-run travel time trends | stop pair + range |
| `tm_aggregate_headways(...)` | Long-run headway trends | stop + range |

Next, check out the **Red Line Analysis** vignette for a full worked example
that goes from raw API data to a polished chart.
