---
title: "Getting Started with transitmattr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with transitmattr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## What is this package?

`transitmattr` lets you pull live and historical data from the
[TransitMatters Data Dashboard API](https://dashboard.transitmatters.org)
directly into R. You can get things like:

- How often trains are coming (headways)
- How long trains sit at stations (dwell times)
- How long it takes to travel between stops (travel times)
- How many people are riding (ridership)
- Where slow zones are affecting service

## Installation

```{r install, eval = FALSE}
# install.packages("pak")  # run this once if you don't have pak
pak::pak("transitmatters/transitmattr")
```

## Loading the package

```{r load}
library(transitmattr)
library(dplyr)
library(ggplot2)
```

## Step 1: Check that the API is working

Before doing anything else, it's good practice to confirm the API is up.

```{r healthcheck, eval = FALSE}
tm_healthcheck()
```

You'll get back something like:

```
$status
[1] "pass"
```

If you see `"pass"` you're good to go.

## Step 2: Understanding what you get back

All `tm_*` functions return a **list** — R's way of bundling different pieces
of data together. Think of it like a backpack: you have to reach in and grab
the thing you want.

```{r list-example, eval = FALSE}
# Stop 70076 is the Park Street northbound platform
result <- tm_headways("2024-01-15", stops = "70076")

# See what's inside:
names(result)

# Grab one piece:
result$headways
```

A helpful trick — `str()` shows you the whole structure at a glance:

```{r str-example, eval = FALSE}
str(result, max.level = 2)
```

## Stop IDs

Several functions (`tm_headways`, `tm_dwells`, `tm_travel_times`) need
**platform stop IDs** rather than station names. Each platform has a separate ID
for each direction. Here are the Red Line trunk stops:

| Station | Northbound (→ Alewife) | Southbound (→ Ashmont/Braintree) |
|---|---|---|
| Alewife | `70061` | `70061` |
| Davis | `70064` | `70063` |
| Porter | `70066` | `70065` |
| Harvard | `70068` | `70067` |
| Central | `70070` | `70069` |
| Kendall/MIT | `70072` | `70071` |
| Charles/MGH | `70074` | `70073` |
| Park Street | `70076` | `70075` |
| Downtown Crossing | `70078` | `70077` |
| South Station | `70080` | `70079` |

You can find stop IDs for all lines via the
[MBTA GTFS feed](https://www.mbta.com/developers/gtfs) or by calling
`tm_facilities()`.

## Step 3: Pull data for a single day

These functions need a date and, for stop-level data, one or more stop IDs.

```{r single-day, eval = FALSE}
# Headways (how often trains arrive) at Park Street northbound
headways <- tm_headways("2024-01-15", stops = "70076")

# Multiple stops at once
headways_multi <- tm_headways("2024-01-15", stops = c("70076", "70064"))

# Dwell times at Park Street northbound
dwells <- tm_dwells("2024-01-15", stops = "70076")

# Travel times between two stops (Park Street → Davis, northbound)
travel <- tm_travel_times("2024-01-15", from_stops = "70076", to_stops = "70064")

# Alerts (service disruptions) — no stop ID needed, but route ID needed
alerts <- tm_alerts("2024-01-15", routes = "Red")
```

## Step 4: Pull data over a date range

Most analysis requires looking at a range of dates. These functions take
`start_date` and `end_date` plus some extra options.

```{r range, eval = FALSE}
# Daily Red Line ridership for January 2024
ridership <- tm_ridership(
  start_date = "2024-01-01",
  end_date   = "2024-01-31",
  line_id    = "line-red"
)

# Daily trip metrics for the Red Line
trips <- tm_trip_metrics(
  start_date = "2024-01-01",
  end_date   = "2024-01-31",
  agg        = "daily",
  line       = "line-red"
)
```

**Available line IDs:**

| Line        | `line_id` / `line`  |
|-------------|---------------------|
| Red Line    | `"line-red"`        |
| Orange Line | `"line-orange"`     |
| Blue Line   | `"line-blue"`       |
| Green Line  | `"line-green"`      |
| Commuter    | `"line-cr"`         |

## Step 5: Turn the result into a data frame

The API gives back a list, but most R analysis tools (like `ggplot2`) prefer a
**data frame**. Here's the easiest way to convert:

```{r to-df, eval = FALSE}
library(dplyr)
library(transitmattr)

# Pull headway data for Park Street northbound
headways_raw <- tm_headways("2024-01-15", stops = "70076")

# Each item is one headway event — bind them into rows
headways_df <- bind_rows(headways_raw)

# Now it looks like a spreadsheet
head(headways_df)
```

> **Tip:** If `bind_rows()` doesn't work, try
> `do.call(rbind, lapply(ridership_raw, as.data.frame))` — it's a bit more
> verbose but handles some tricky list shapes.

## Step 6: Make a quick plot

Once you have a data frame you can plot with `ggplot2`:

```{r plot, eval = FALSE}
library(ggplot2)
library(dplyr)
library(transitmattr)

ridership_df <- bind_rows(
  tm_ridership("2024-01-01", "2024-01-31", line_id = "line-red")
)

ggplot(ridership_df, aes(x = as.Date(date), y = count)) +
  geom_line(color = "#DA291C") +   # Red Line color
  labs(
    title = "Red Line Daily Ridership — January 2024",
    x     = "Date",
    y     = "Riders"
  ) +
  theme_minimal()
```

## Recap: the functions at a glance

| Function | What it returns | Needs |
|---|---|---|
| `tm_healthcheck()` | API status | nothing |
| `tm_facilities()` | All MBTA stations | nothing |
| `tm_alerts(date)` | Service alerts | optional date |
| `tm_headways(date, stops)` | Train frequency | date + stop IDs |
| `tm_dwells(date, stops)` | Station dwell times | date + stop IDs |
| `tm_travel_times(date, from, to)` | Stop-to-stop travel | date + stop ID pair |
| `tm_ridership(start, end)` | Ridership counts | date range |
| `tm_trip_metrics(start, end, agg, line)` | Trip performance | date range + line |
| `tm_line_delays(start, end, line)` | Delay summaries | date range + line |
| `tm_speed_restrictions(line_id, date)` | Slow zones | line + date |
| `tm_aggregate_travel_times(...)` | Long-run travel time trends | stop pair + range |
| `tm_aggregate_headways(...)` | Long-run headway trends | stop + range |

Next, check out the **Red Line Analysis** vignette for a full worked example
that goes from raw API data to a polished chart.
