---
title: "Red Line Analysis: A Worked Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Red Line Analysis: A Worked Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE  # set to TRUE locally to actually hit the API
)
```

This vignette walks through a realistic analysis of Red Line performance in
January 2024. By the end you'll have three charts:

1. Daily ridership
2. Average travel time from Park Street → Davis
3. Where slow zones were on a specific day

## Setup

```{r load, eval = FALSE}
library(transitmattr)
library(dplyr)
library(ggplot2)
```

> **New to these packages?**
> `dplyr` is for cleaning and reshaping data; `ggplot2` is for charts.
> Install them with `install.packages(c("dplyr", "ggplot2"))` if needed.

## 1. Daily ridership

### Pull the data

Build a Red Line query, set a date range, and call `$ridership()`. The result
is already a `data.frame` — no `bind_rows()` needed.

```{r ridership-pull, eval = FALSE}
ridership_df <- tm_rt_query("Red")$date_range("2024-01-01", "2024-01-31")$ridership()
head(ridership_df)
```

### Plot it

```{r ridership-plot, eval = FALSE}
ggplot(ridership_df, aes(x = as.Date(date), y = count)) +
  geom_col(fill = "#DA291C") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Red Line Daily Ridership",
    subtitle = "January 2024",
    x        = NULL,
    y        = "Riders"
  ) +
  theme_minimal()
```

> **What to notice:** Weekends typically show lower ridership. If a weekday is
> unusually low, there may have been a service disruption — check
> `tm_alerts()` for that date.

## 2. Travel time from Park Street to Davis

This uses the **aggregate** endpoint, which gives you long-run daily averages
rather than individual trips.

### Pull the data

Specify from/to by station name. The library handles stop ID resolution
internally — no need to look up `"70076"` or `"70064"`.

```{r travel-pull, eval = FALSE}
travel_df <- tm_rt_query("Red")$from_stop("Park Street")$to_stop("Davis")$date_range("2024-01-01", "2024-01-31")$aggregate_travel_times()
head(travel_df)
```

### Clean and plot

```{r travel-plot, eval = FALSE}
travel_df <- travel_df |>
  mutate(
    date       = as.Date(service_date),
    travel_min = mean / 60
  )

ggplot(travel_df, aes(x = date, y = travel_min)) +
  geom_line(color = "#DA291C") +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed",
              color = "grey40") +
  labs(
    title    = "Average Travel Time: Park Street → Davis",
    subtitle = "January 2024 · Red Line",
    x        = NULL,
    y        = "Minutes"
  ) +
  theme_minimal()
```

> **Dashed line** is a smooth trend. If travel times creep up over time it can
> signal slow zones accumulating — which leads us to section 3.

## 3. Slow zones on a specific day

Speed restrictions ("slow zones") are places where trains must go slower than
normal, usually because of track issues. Unlike most endpoints, this one
returns a **list** (not a data.frame) because its structure is more complex.

```{r slow-pull, eval = FALSE}
slow_raw <- tm_rt_query("Red")$speed_restrictions("2024-01-15")
```

```{r slow-df, eval = FALSE}
if (!isTRUE(slow_raw$available)) {
  message("No slow zone data available for this date.")
} else {
  slow_df <- bind_rows(slow_raw$zones)
  slow_df
}
```

```{r slow-plot, eval = FALSE}
ggplot(slow_df, aes(x = reorder(description, speedMph), y = speedMph, fill = speedMph)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "#DA291C", high = "#f9c0c0") +
  labs(
    title    = "Red Line Slow Zones — January 15, 2024",
    subtitle = "Lower mph = more restricted",
    x        = NULL,
    y        = "Speed limit (mph)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

> **Note:** If `slow_raw$available` is `FALSE`, there were no active slow zones
> that day. Try a different date or check during a period with known track work.

## 4. Putting it all together

Here's the pattern you'll use over and over with `transitmattr`:

```r
tm_rt_query("Red")             # pick a line
  $stop("Harvard")             # set a station
  $date("2024-01-15")          # set a date
  $headways()                  # fetch — returns a data.frame
```

Once you're comfortable with it, try exploring other methods:

- `$line_delays()` — delay minutes caused by alerts
- `$trip_metrics("daily")` — on-time performance per trip
- `$aggregate_headways()` — how often trains come, over time
- `$service_hours("weekly")` — scheduled vs. delivered service hours

## Troubleshooting

| Problem | What to try |
|---|---|
| "requires a stop" error | Call `$stop("Name")` before the terminal method |
| "requires a date range" error | Call `$date_range(start, end)` before the terminal method |
| All values are `NULL` | The API may have no data for that date; try a nearby weekday |
| The API call fails | Run `tm_healthcheck()` to verify the API is up |
