---
title: "Red Line Analysis: A Worked Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Red Line Analysis: A Worked Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE   # set to TRUE locally to actually hit the API
)
```

This vignette walks through a realistic analysis of Red Line performance in
January 2024. By the end you'll have three charts:

1. Daily ridership
2. Average travel time from Park Street → Davis
3. Where slow zones were on a specific day

## Setup

```{r load}
library(transitmattr)
library(dplyr)
library(ggplot2)
```

> **New to these packages?**
> `dplyr` is for cleaning and reshaping data; `ggplot2` is for charts.
> Install them with `install.packages(c("dplyr", "ggplot2"))` if needed.

## 1. Daily ridership

### Pull the data

```{r ridership-pull}
ridership_raw <- tm_ridership(
  start_date = "2024-01-01",
  end_date   = "2024-01-31",
  line_id    = "Red"
)
```

`ridership_raw` is a **list** where each item represents one day's record.
Let's flatten it into a data frame:

```{r ridership-df}
library(dplyr)
ridership_df <- bind_rows(ridership_raw)
head(ridership_df)
```

### Explore the columns

```{r ridership-cols}
glimpse(ridership_df)
```

### Plot it

```{r ridership-plot}
library(ggplot2)
ggplot(ridership_df, aes(x = as.Date(date), y = count)) +
  geom_col(fill = "#DA291C") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Red Line Daily Ridership",
    subtitle = "January 2024",
    x = NULL,
    y = "Riders"
  ) +
  theme_minimal()
```

> **What to notice:** Weekends typically show lower ridership. If a day is
> unusually low on a weekday, there may have been a service disruption — check
> `tm_alerts()` for that date.

## 2. Travel time from Park Street to Davis

This uses the **aggregate** endpoint, which gives you long-run averages
rather than individual trips.

### Finding stop IDs

Travel time endpoints need **platform stop IDs**, not parent station IDs. Each
station has separate IDs for each direction. Use `tm_stops()` to look them up:

```{r stops-lookup, eval = FALSE}
red_stops <- tm_stops("Red")

# Stations are in red_stops$stations — explore one
str(red_stops$stations[[1]])
```

> **Tip:** Northbound (toward Alewife) stop IDs are even numbers; southbound
> (toward Ashmont/Braintree) are odd. For example, Park Street northbound is
> `70076` and southbound is `70075`.

### Pull the data

```{r travel-pull}
# Northbound: Park Street (70076) → Davis (70064)
travel_raw <- tm_aggregate_travel_times2(
  from_stop  = "70076",
  to_stop    = "70064",
  start_date = "2024-01-01",
  end_date   = "2024-01-31"
)
```

### Clean it up

```{r travel-df}
library(dplyr)
travel_df <- bind_rows(travel_raw$by_date)

# Travel times come back in seconds — convert to minutes
travel_df <- travel_df |>
  mutate(
    date       = as.Date(service_date),
    travel_min = mean / 60
  )

head(travel_df)
```

### Plot it

```{r travel-plot}
library(ggplot2)
ggplot(travel_df, aes(x = date, y = travel_min)) +
  geom_line(color = "#DA291C") +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed",
              color = "grey40") +
  labs(
    title    = "Average Travel Time: Park Street → Davis",
    subtitle = "January 2024 · Red Line",
    x        = NULL,
    y        = "Minutes"
  ) +
  theme_minimal()
```

> **Dashed line** is a smooth trend. If travel times creep up over time it can
> signal slow zones accumulating — which leads us to section 3.

## 3. Slow zones on a specific day

Speed restrictions ("slow zones") are places where trains must go slower than
normal, usually because of track issues.

```{r slow-pull}
slow_raw <- tm_speed_restrictions("Red", "2024-01-15")
```

```{r slow-df}
library(dplyr)
if (!isTRUE(slow_raw$available)) {
  message("No slow zone data available for this date.")
} else {
  slow_df <- bind_rows(slow_raw$zones)
  glimpse(slow_df)
}
```

```{r slow-plot}
library(ggplot2)
# Each row is a slow zone segment; visualize by location and speed limit
ggplot(slow_df, aes(x = reorder(description, speedMph), y = speedMph, fill = speedMph)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "#DA291C", high = "#f9c0c0") +
  labs(
    title    = "Red Line Slow Zones — January 15, 2024",
    subtitle = "Lower mph = more restricted",
    x        = NULL,
    y        = "Speed limit (mph)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

> **Note:** If `slow_raw$available` is `FALSE`, there were no active slow zones
> that day. Try a different date or check during a period with known track work.

## 4. Putting it all together

Here's a pattern you'll use over and over: **pull → bind_rows → mutate →
ggplot**. Once you're comfortable with it, try exploring:

- `tm_routes()` — see all available lines before picking one
- `tm_stops("Orange")` — look up stop IDs for a different line
- `tm_aggregate_headways()` — how often trains come, over time
- `tm_line_delays()` — delay minutes caused by alerts
- `tm_trip_metrics()` — on-time performance per trip

## Troubleshooting

| Problem | What to try |
|---|---|
| `bind_rows()` gives an error | Use `str(raw_result)` to see the exact structure, then drill in with `$` |
| All values are `NULL` | The API may have no data for that date; try a nearby weekday |
| Columns have unexpected names | Run `names(df)` and adjust your `mutate()` references |
| The API call fails | Run `tm_healthcheck()` to verify the API is up |
